Longitudinal Principal Manifold Estimation

### 2023-05-09 paper/main.tex
Add Kun comments

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\title{Longitudinal Manifold Estimation}
\author[1]{Robert Zielinski, Kun Meng, and Ani Eloyan}

@@

The remainder of this article is laid out as follows: In Section 2, we introduce the principal manifold framework proposed in \cite{mengPrincipalManifoldEstimation2021}. Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this. Section 4 demonstrates the performance of this approach on simulated data. In Section 5, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease. The article concludes with a discussion of the method's contributions in Section 6.

@@

The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of the data. \cite{hastiePrincipalCurves1989} defines principal curves as follows:

@@

To completely define principal manifolds, \cite{mengPrincipalManifoldEstimation2021} let $\|Pf\|_{\mathcal{H}} = \|f''(r)\|_{\mathbb{R}^{D}}^2$, which provides a greater penalization of roughness than using the first derivative. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
\begin{equation}
  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
\end{equation}
referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:
\begin{definition}
  \label{def:principal_manifolds}
  Let $X$ be a random $D$-vector associated with the probability measure or density function $\mathbb{P}$ such that $X$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments. Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
  \begin{align}
    \mathcal{K}_{\lambda, \mathbb{P}}(f, g) &= E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 \label{eq:5} \\
    \mathcal{K}_{\lambda, \mathbb{P}}(f) &= \mathcal{K}_{\lambda, \mathbb{P}}(f, f) \label{eq:6}
  \end{align}

  Then a manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $X$ with the tuning parameter $\lambda$ if 
  \begin{equation}
    f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \ \text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}. \label{eq:7}
  \end{equation}
\end{definition}

In this setting, we can see that $\pi_{f_{\lambda}^{*}}(X)$ maps the $D$-vector $X$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.

To estimate principal manifolds under this definition, \cite{mengPrincipalManifoldEstimation2021} propose the Principal Manifold Estimation (PME) algorithm. This algorithm has three main steps: data reduction, fitting, and tuning. In the reduction step, the HDMDE algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$. The number of clusters is chosen using an iterative hypothesis testing approach. For each given value of $\lambda$, the fitting stage of the algorithm uses this reduced dataset to find estimated function 
\begin{equation}
  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f), \label{eq:8}
\end{equation}
where 
\begin{equation}
  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 \label{eq:9}
\end{equation}
approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses the optimal value of $\lambda$, $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
\begin{equation}
  \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2 \label{eq:10}
\end{equation}
is an empirical estimate of the MSD using the full dataset.

\cite{mengPrincipalManifoldEstimation2021} show that the optimal function will be of spline form, meaning that when $d = 1$, the function will be a cubic smoothing spline, while the function is a thin plate spline when $d = 2$. Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in \cite{mengPrincipalManifoldEstimation2021}.

@@

We now introduce the proposed longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as \cite{mengPrincipalManifoldEstimation2021}. Let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points. As in the setting described above, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. However, in this updated case, the function determining the manifold may change over time. Note that we assume changes over time to be smooth. Therefore, we seek a function $f^{*}$ that minimizes $\mathcal{D}_X(f^{*})$ while penalizing both the roughness of the function over the time-independent shape and the changes in the function over time, expressed as 
\begin{equation}
  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(f, g) = E\|X - f(\pi_g(X))\|_{\mathbb{R}^{D}}^2 + \sum_{t=1}^{T} \lambda_t \|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 + \gamma\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^d)}^2 \label{eq:11}
\end{equation}

To solve this problem, we propose an extension of the method developed in \cite{mengPrincipalManifoldEstimation2021}. The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Details of this approach are described below.

\subsection*{Initialization}
We begin by considering $D$-dimensional observations $\left\{y_{it}\right\}_{i=1, t=1}^{I_t, T}$. Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\left\{\hat{\theta}_{j, t}\right\}$. HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure. The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

The PME algorithm uses ISOMAP (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another). Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed. Instead of using ISOMAP on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one ISOMAP run to reach a parameterization for the centers across all time points simultaneously. Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time.

\subsection*{Fitting}
Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point. Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run. This results in an estimated principal manifold for each time point, $f_t$, and an optimal tuning parameter $\lambda_t^*$. The remainder of the fitting stage involves estimating a thin plate spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$.

At each time point, the estimated principal manifold has the functional form of a smoothing spline or thin plate spline, depending on the intrinsic dimension $d$. As detailed by \cite{greenSilverman1994}, smoothing splines, and thin plate splines as their higher dimensional analogues, may be considered to have knots at the points being used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fit must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\mathbf{r}_i\right\}_{i=1}^{N}$ ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions for each time point $f_t$. This yields a set of outcomes, $\left\{X_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\mathbf{s}_t^*$ and $\mathbf{\alpha}_t^*$ can then be calculated by minimizing
\begin{equation}
  \mathcal{L}(f_t) = (\mathbf{X_t} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t)^{T}(\mathbf{X} - E\mathbf{s}_t - R^{T}\mathbf{\alpha}_t) + \gamma\mathbf{s}_t^{T}E\mathbf{s}_t, \label{eq:12}
\end{equation}
where matrix $E$ is defined by $E_{ij} = \eta_{d}(\|\mathbf{r}_i - \mathbf{r}_j\|)$ and matrix $R$ has $R_{ij} = \phi_i(\mathbf{r}_j)$, where $\phi_i(\cdot)$ is a linear basis function. This minimization is conducted by solving
\begin{equation}
  \left(
    \begin{array}{cc}
      E + \lambda_t^* I & R^{T} \\
      R^{T} & 0
    \end{array}
  \right)\left(
    \begin{array}{c}
      \mathbf{s}_t \\
      \mathbf{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \mathbf{X}_t \\
      0
    \end{array}
  \right), \label{eq:13}
\end{equation}
yielding function
\begin{equation}
  f_t^*(r) = \sum_{j=1}^{N}s_{jt}^* \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{kt}^*\phi(r)_k. \label{eq:14}
\end{equation}

>>>>

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\T}{\intercal}
\newcommand{\commentout}[1]{}

\newcommand{\kmedit}[1]{{\color{purple}  #1}}
\newcommand{\meng}[1]{{\color{purple} \sf $\clubsuit\clubsuit\clubsuit$ Kun Meng: [#1]}}
\newcommand{\Meng}[1]{\margMa{(Kun Meng) #1}}

@@

\title{Longitudinal Principal Manifold Estimation}
\author[1]{Robert Zielinski, Kun Meng, and Ani Eloyan}

@@

The remainder of this article is laid out as follows: In Section 2, we review the principal manifold framework proposed in \cite{mengPrincipalManifoldEstimation2021}. Section 3 discusses how this approach can be adapted for longitudinal settings, and shows the algorithm being used to accomplish this. Section 4 demonstrates the performance of this approach on simulated data. In Section 5, we apply this method to longitudinal data of the hippocampus in individuals with Alzheimer's disease. The article concludes with a discussion of the method's contributions in Section 6.

@@

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of the data. However, the PS framework in \cite{hastiePrincipalCurves1989} is flawed by the model complexity selection, which \cite{duchamp1996extremal} pointed out. Motivated by the penalization approaches in \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework and algorithm. Since our LPME is based on the PME, we review the PME in this section for the audience's convenience. The following concepts and notations are necessary for both the PME and our proposed LPME, and they will be used throughout this paper
\begin{itemize}
\item $d$ and $D$ are two positive integers denoting dimensions, and $d<D$; \meng{Add some explanations of the dimensionalities $d$ and $D$.}
\item for any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$;
    \item $\mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D}) = \left\{f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D}): \lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty\right\}$, where $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ denotes the collection of continuous vector-valued functions mapping from $\mathbb{R}^d$ to $\mathbb{R}^D$; 
    \item for every $f\in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, the image of $f$ is referred to as the $d$-dimensional manifold $M_f^d$ determined by $f$, i.e., $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$;
    \item For each $f\in \mathcal{C}_{\infty}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and each point $x\in\mathbb{R}^D$, $\pi_f(x)$ is a point in $\mathbb{R}^d$ and denotes the ``projection index" of $x$; roughly, $\pi_f(x)$ is a point in $d$-dimensional space such that $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$; $\pi_f(x)$ is a generalization of the counterpart in \cite{hastiePrincipalCurves1989}, and its details can be found in Section 2 of \cite{mengPrincipalManifoldEstimation2021};
    \item suppose $u(r_1,\ldots,r_d)$ is a scalar-valued function defined on $\mathbb{R}^d$; $\nabla^{\otimes 2}u=\left(\frac{\partial^2 u}{\partial r_i \partial r_j}\right)_{1\le i,j\le d}$ denotes the Hessian matrix of $u$, and $\|\nabla^{\otimes 2}u\|_{\mathbb{R}^{d \times d}}$ is the function $r \to \|\nabla^{\otimes 2} u(r)\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2\right)^{\frac{1}{2}}$; in addition, $\|\nabla^{\otimes 2} u\|_{L^2(\mathbb{R}^{d})}^2 = \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2 u}{\partial r_i \partial r_j}(r)\right|^2dr $, and 
    \begin{align*}
        \nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\},
    \end{align*}
    where $\mathcal{D}'$ denotes the collection of distributions (also known as generalized functions, see \cite{duchon1977splines} and Chapter 6 of \cite{rudin1991functional});
    \item $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$ denotes the collection of vector-valued functions $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ whose components are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d})$; besides, $\mathcal{C}_\infty\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)=\mathcal{C}_\infty(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)\bigcap\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\rightarrow\mathbb{R}^D)$; the following functional of $f(r)=\left(f_1(r),\ldots,f_D(r)\right)^\T$ measures the curvature of the manifold $M_f^d$ and will be used throughout this paper
    \begin{equation}
  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \|\nabla^{\otimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2  = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
\end{equation}
\end{itemize}
With the preparations above, the definition of principal manifolds proposed in \cite{mengPrincipalManifoldEstimation2021} is presented as follows
\commentout{
\cite{hastiePrincipalCurves1989} defines principal curves as follows:

@@

To completely define principal manifolds, \cite{mengPrincipalManifoldEstimation2021} let $\|Pf\|_{\mathcal{H}} = \|f''(r)\|_{\mathbb{R}^{D}}^2$, which provides a greater penalization of roughness than using the first derivative. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, they generalize this term as
\begin{equation}
  \|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr, \label{eq:4}
\end{equation}
referred to as the total squared curvature of a manifold. With this penalty term set, principal manifolds are then defined as follows:}
\begin{definition}
  \label{def:principal_manifolds}
  Let $\boldsymbol{X}$ be a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support $\text{supp}(\mathbb{P})$ and finite second moments. Let $f, g \in \mathcal{C}_{\infty}\cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D})$ and $\lambda \in [0, \infty]$, and define the following functionals:
  \begin{align}\label{eq:5}
    \begin{aligned}
&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\
&\mathcal{K}_{\lambda, \mathbb{P}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f).
    \end{aligned}
  \end{align}
A manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ is called a principal manifold for $\boldsymbol{X}$ with the tuning parameter $\lambda$ if 
  \begin{equation}\label{eq:7}
    \begin{aligned}
        & f_{\lambda}^{*} = \argmin_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f), \\ 
    &\text{where} \ \mathcal{F}(\mathbb{P}) = \left\{f \in \mathcal{C}_{\infty} \cap \nabla^{-\otimes 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}. 
    \end{aligned}
  \end{equation}
\end{definition}
\meng{Add some explanations for the terms in Eq.~\eqref{eq:5}.}
In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space.

To estimate principal manifolds under this definition, \cite{mengPrincipalManifoldEstimation2021} propose the Principal Manifold Estimation (PME) algorithm. This algorithm has three main steps: data reduction, fitting, and tuning (see Figure 2 of \cite{mengPrincipalManifoldEstimation2021}). \meng{I would suggest presenting Figure 2 of \cite{mengPrincipalManifoldEstimation2021} in the present paper.}
\begin{enumerate}
    \item In the data reduction step, the high-dimensional mixture density estimation (HDMDE) algorithm uses $k$-means clustering to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by cluster centers $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$. The number $N$ of clusters is chosen using an iterative hypothesis testing approach, and $N$ tends to be much smaller than $I$ (see the simulation study in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). 
    \item For each given value of $\lambda$, the fitting stage of the PME algorithm uses this reduced data set $\left\{\mu_j\right\}_{j=1}^{N}$ to find estimated function 
\begin{equation}
  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f), \label{eq:8}
\end{equation}
where $\mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2$ approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. 
\item The tuning stage chooses the optimal value of the parameter $\lambda$ by $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
\begin{equation}
  \mathcal{D}\left(\hat{f}_\lambda\right) = \frac{1}{I}\sum_{i=1}^{I}\left\|x_i - \hat{f}_\lambda\left(\pi_{\hat{f}_\lambda}(x_i)\right)\right\|_{\mathbb{R}^{D}}^2 \label{eq:10}
\end{equation}
is an empirical estimate of the MSD using the full dataset.
\end{enumerate}

Using results in \cite{duchon1977splines} and \cite{wahba1990}, \cite{mengPrincipalManifoldEstimation2021} show that the optimal function will be of a spline form. Specifically, when $d = 1$, the function will be a cubic smoothing spline, while the function is a thin plate spline when $d = 2$. Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in \cite{mengPrincipalManifoldEstimation2021}.

\meng{I commented out many literature review discussions about \cite{hastiePrincipalCurves1989} and \cite{smolaRegularizedPrincipalManifolds2001} to make this section more concise. In addition, if the formulas above are not referred to in our discussions, they shouldn't be given labels.}

@@

We now propose longitudinal principal manifold estimation (LPME) algorithm, following a similar setting as \cite{mengPrincipalManifoldEstimation2021}.

Let $\left\{x_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ $D$-dimensional observations for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. However, in this updated case, the function determining the manifold may change over time.

\subsection{Definition of LPM}

Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively. In addition, we assume $F\in\mathcal{C}_\infty\cap\nabla^{-\otimes2}L^2(\mathbb{R}^d\times\mathbb{R}\rightarrow\mathbb{R}^D)$. For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold at the time point. For given tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, we plan to minimize the following functional
\begin{equation}
  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) = \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt. \label{eq:11}
\end{equation}
\meng{Combine Eq.~\eqref{eq:5} and the functional above.}

\meng{Add the interpretation of the terms above. This interpretation presents the key idea of this paper. I need to address it carefully. One thing I would strongly suggest is to explain why we chose to deal with space and time separately.}

\meng{Mentioned the functional above in all the four steps of LPME algorithm --- to emphasize that we are going to minimize the functional above.}

\commentout{Note that we assume changes over time to be smooth. Therefore, we seek a function $f^{*}$ that minimizes $\mathcal{D}_X(f^{*})$ while penalizing both the roughness of the function over the time-independent shape and the changes in the function over time, expressed as }

\subsection{Alogirthm}

To solve this problem, we propose an extension of the PME algorithm developed in \cite{mengPrincipalManifoldEstimation2021}. The output of the PME algorithm run used for the data at an individual time point is a spline function mapping from a low-dimensional paramterization to a high-dimensional space. In the LPME algorithm, we use the coefficients of PME functions fit at each time point as the outcomes in a thin plate spline, thus smoothing over the coefficients with respect to time. This yields a multistage model that, for a provided time point and set of parameters, obtains a set of spline coefficients associated with the time value, then returns the high-dimensional output of the spline function associated with the parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. Details of this approach are described below.

\meng{I would suggest making a flowchart similar to Figure 2 of \cite{mengPrincipalManifoldEstimation2021}.}

\subsection*{Data Reduction}
We begin by considering $D$-dimensional observations $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$. Like PME, the LPME algorithm begins with a data reduction step, using the HDMDE method detailed in \cite{mengPrincipalManifoldEstimation2021} (Section 5.3 therein) to estimate a reduced set of cluster centers, $\left\{\mu_{j, t}\right\}_{j=1, t=1}^{N_t, T}$, along with corresponding weights $\{\hat{\theta}_{j, t}\}_{j=1, t=1}^{N_t, T}$. HDMDE is run separately for each individual time point to ensure the time dimension is preserved, while the number of clusters being estimated is allowed to vary for each time point to accommodate potential changes in the shape of the underlying structure.

This step reduces the size $\sum_{t=1}^T I_t$ of data $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ to the size $\sum_{t=1}^T N_t$, where each $N_t$ tends to be much smaller than the corresponding $I_t$ (see the simulation study presented in Figure 3(c) of \cite{mengPrincipalManifoldEstimation2021}). The centers estimated in this step are then used to reach an initial low-dimensional parameterization.

\subsection*{Initialization}
The PME algorithm uses Isomap (\cite{tenenbaumGlobalGeometricFramework2000}) to develop an initial parameterization, which the algorithm proceeds to iteratively improve. However, using this process several times over similarly shaped structures is likely to result in inconsistent parameterization directions (i.e. low values in the parameter space may be associated with low values in the embedding space for one run, while low values in the parameter space may be associated with high values in the embedding space in another) \meng{I would suggest writing more explanations for the ``inconsistent parameterization directions." Some visual illustrations are necessary.}. Reaching a successful fit in the LPME algorithm requires consistent parameterizations across time points, so a different approach to generating an initial parameterization is needed. \meng{This paragraph should be devoted to the explanation for ``inconsistent parameterization directions." The comparison can be presented using visualizations.}

Instead of using Isomap on the data for each time point separately, the LMPE approach collapses the time dimension in the cluster centers of all time steps, then uses one Isomap run to reach a parameterization for the centers across all time points simultaneously. Assuming relative stability in the structure shape over time, this process ensures that the initial parameterization remains consistent over time. \meng{I would suggest presenting simulation studies showing the ``consistency over time." Specifically, we may compare the difference between time-wise and over-time initializations via simulations.}

\subsection*{Fitting}
Once the initial parameterization for each cluster center is obtained, the model fitting process begins by using this parameterization as the initialization for runs of the PME algorithm at each time point. Specifically, for each time point, the cluster centers associated with the given time value, their weights, and the corresponding parameterizations for these points are used to initialize a PME run. This results in an estimated principal manifold $\widehat{f_t}$ for each time point $t$ and an optimal tuning parameter $\lambda_t^*$, which deals with the first two terms of the $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ defined in Eq.~\eqref{eq:11}.

The remainder of the fitting stage deals with the third term of $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$, i.e., the curvature in the time dimension (see Eq.~\eqref{eq:11}). Specifically, the remainder of this stage involves estimating a spline function mapping from the time dimension to the coefficient space that smooths over the coefficients for these principal manifolds, given a tuning parameter $\gamma$. \meng{The term ``coefficient space" is clear to me, but it might be confusing to other readers who are unfair with the details of PME.}

At each time point, the estimated principal manifold has a spline form. As detailed by \cite{greenSilverman1994}, splines can be considered to have knots at the points being used to fit them. This means that, in order for the coefficients of the manifolds estimated at each time point to be comparable, the parameter values used as inputs when the functions were fitted must be identical. Hence, to obtain coefficients that can be compared across time points, a grid of values $\left\{\boldsymbol{r}_i\right\}_{i=1}^{N}$ ($\subset\mathbb{R}^d$) ranging from the minimum to the maximum of the parameters in each of the $d$ dimensions is input into the estimated embedding functions $\widehat{f_t}$ for each time point  \meng{What is the $N$ herein?}. This yields a set of outcomes, $\left\{X_{i,t}\right\}_{i=1, t = 1}^{N, T}$, corresponding to same parameter values at each time point, i.e., $X_{i,t}=\widehat{f_t}(\boldsymbol{r}_i)$ for all $i$ and $t$. Following the discussion in Chapter 7 of \cite{greenSilverman1994}, for each time point $t$, comparable spline coefficients $\boldsymbol{s}_t^*$ and $\boldsymbol{\alpha}_t^*$ can then be calculated by minimizing $\mathcal{L}(f_t)$, where
\begin{itemize}
    \item $\mathcal{L}(f_t) = (\boldsymbol{X}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t)^\T(\boldsymbol{X}_t - \boldsymbol{E}\boldsymbol{s}_t - \boldsymbol{R}^\T\boldsymbol{\alpha}_t) + \gamma\boldsymbol{s}_t^\T \boldsymbol{E}\boldsymbol{s}_t$;
    \item \meng{specify the function $\eta_{4-d}$};
    \item matrix $\boldsymbol{E}=(E_{ij})_{1\le i,j\le N}$ is defined by $E_{ij} = \eta_{d}(\|\boldsymbol{r}_i - \boldsymbol{r}_j\|)$ \meng{$\eta_d$ or $\eta_{4-d}$?};
    \item functions $\phi_1,\ldots,\phi_{d+1}$ form a basis of the linear space of polynomials on $\mathbb{R}^d$ with degrees $\le1$;
    \item matrix $\boldsymbol{R}=(R_{ij})_{1\le i\le d+1,1\le j\le N}$ is defined by has $R_{ij} = \phi_i(\boldsymbol{r}_j)$.
\end{itemize}
This minimization is conducted by solving
\begin{equation}
  \left(
    \begin{array}{cc}
      \boldsymbol{E} + \lambda_t^* \boldsymbol{I} & \boldsymbol{R}^\T \\
      \boldsymbol{R} & \boldsymbol{0}
    \end{array}
  \right)\left(
    \begin{array}{c}
      \boldsymbol{s}_t \\
      \boldsymbol{\alpha}_t
    \end{array}
  \right) = \left(
    \begin{array}{c}
      \boldsymbol{X}_t \\
      \boldsymbol{0}
    \end{array}
  \right). \label{eq:13}
\end{equation}
The solution to Eq.~\eqref{eq:13} , denoted by $\boldsymbol{\alpha}^*_t=(\alpha^*_{1,t},\ldots,\alpha^*_{d+1,t})^\T$ and $\boldsymbol{s}^*_t=(s^*_{1,t},\ldots,s^*_{N,t})^\T$, yields function
\begin{equation}
  f_t^*(r) = \sum_{j=1}^{N}s_{j,t}^* \cdot  \eta_{4-d}\left(\|r - r_j^*\|\right) + \sum_{k=1}^{d + 1}\alpha_{k,t}^* \cdot \phi_k(r). \label{eq:14}
\end{equation}
